### 2018-02-10 submission/manuscript.Rmd
update and improve discussion

Our study identifies clear differences in diversity, both at the community level and for individual genera, present in patients with and without CRC [Figure 1-3]. Although there was a step-wise decrease in diversity as disease progressed from control to adenoma to carcinoma, this did not translate into large effect sizes for the OR of tumors. Even though mouth-associated genera increased the OR of having a carcinoma, they did not consistently increase the RR of having an adenoma. Additionally, our observations suggest that by combining mouth-associated and CRC protective microbes we can classify either adenoma or carcinoma stage of disease as well as models that use the full community.

The data presented herein support the importance of select genera for carcinoma, but not necessarily adenoma, formation. The results that we have presented show that both the genera and OTU select and full models, for the carcinoma stage of CRC, had similar AUCs [Figure 4 & 5]. This suggests that an interplay between a select number of potentially protective and exacerbating microbes within the GI community is crucial for carcinoma formation. Importantly, it suggests that there may be key members of the GI community that might be studied further to potentially reduce the risk of carcinoma. Conversely, using the present data, it is clear that new approaches may be needed to identify members of the community associated with adenoma stage of disease. Regardless of sample type and whether a full or select model was used, our Random Forest models consistently performed poorly. Yet, the step-wise decrease in diversity suggests that the adenoma-associated community is not normal but has changed subtly [Figure 1]. This change in diversity, at this early stage of disease, could be focal to the adenoma itself. One possible explanation is that how the host interacts with these subtle changes at early stages of the disease is what leads to a thoroughly dysfunctional community that is supportive of CRC genesis.

Within stool, common GI microbes were most consistently present in the top 10 genera or OTUs across studies [Figure 6]. Changes in *Bacteroides*, *Ruminococcaceae*, *Ruminococcus*, and *Roseburia* were consistently found to be in the top 10 most important variables across the different studies for both adenoma and carcinoma [Figure 6]. These data suggest that whether the non-resident bacterium is *Fusobacteria* or *Peptostreptococcus* may not be as important as how these bacteria interact with the changing resident community. Based on these observations, it is possible to hypothesize that small changes in community structure lead to new niches in which any one of the mouth-associated genera can gain a foothold, exacerbating the initial changes in community and facilitating the transition from adenoma to carcinoma stage of disease.

@@

One important caveat to this study is that even though genera associated with certain species such as *Bacteroides fragilis* and *Streptococcus gallolyticus* subsp. *gallolyticus* were not identified, it does not necessarily mean that these specific species are not important in human CRC [@bfrag_Boleij_2014; @strep_Aymeric_2017]. Since we are limited in our aggregation of the data to the genus level, it is not possible to clearly delineate which species are contributing to overall disease progression. Our observations are not inconsistent with the previous literature on either *Bacteroides fragilis* or *Streptococcus gallolyticus* subsp. *gallolyticus*. As an example, the stool-based full community models consistently identified the genus *Bacteroides*, as well as OTUs that classified as *Bacteroides*, to be important model components across studies. This suggests that even though *Bacteroides* may not increase the RR of CRC and may not vary in relative abundance, like *Fusobacterium*, it is still important in CRC. Additionally, *Streptococcus gallolyticus* subsp. *gallolyticus* is a mouth-associated microbe, and the results from this study suggest that regardless of sample type, mouth-associated genera are commonly associated with an increased RR for both adenoma and carcinoma stage of disease.

The associations between the microbiota and adenoma stage of disease are inconclusive, in part, because many studies may not be powered effectively to observe small effect sizes. None of the studies analyzed were properly powered to detect a 10% or lower change between cases and controls. The results within our meta-analysis suggest that a small effect size may well be the scope in which differences consistently occur between controls and adenoma stage of disease. Future studies investigating adenoma stage and the microbiota need to take power into consideration to reproducibly study whether the microbiota contributes to polyp formation. In contrast to adenoma stage of disease, our observations suggest that most studies analyzed have sufficient power to detect many changes in the carcinoma-associated microbiota because of large effect size differences between cases and controls [Figure 7].

>>>>

Our study identifies clear differences in diversity, both at the community level and for individual genera, present in patients with tumors versus controls [Figure 1-3]. Although there was a step-wise decrease in diversity as disease progressed from control to adenoma to carcinoma, this did not translate into large effect size increases in OR for either adenoma or carcinoma tumors. Even though mouth-associated genera increased individaul's OR of having a carcinoma, they did not consistently increase the OR of having an adenoma. Additionally, our observations suggest that by combining the top 5 most protective and detrimental microbes we can classify indviduals with either adenoma or carcinoma as well as models that use the full community.

The data presented herein support the importance of select genera for carcinoma, but not necessarily adenoma, tumor formation. The results that we have presented show that both the genera- and OTU-based select and full models, for indviduals with carcinoma, had similar AUCs [Figure 4 & 5]. This suggests that an interplay between a select number of potentially protective and exacerbating microbes within the GI community could be crucial for carcinoma formation. Importantly, it suggests that there may be key members of the GI community that should be studied further to potentially help reduce the risk of carcinoma tumor formation. Conversely, using the present data, it is clear that new approaches may be needed to identify members of the community associated with adenoma turmors. Regardless of sample type and whether a full or select model was used, our Random Forest models consistently performed poorly. Yet, the step-wise decrease in diversity suggests that the adenoma-associated community is not normal but has changed subtly [Figure 1]. This change in diversity, at this early stage of disease, could be focal to the adenoma itself. Additionally, how the host interacts with these subtle changes at early stages of the disease could be what leads to a thoroughly dysfunctional community that is supportive of tumorgenesis.

Within stool, common GI microbes were most consistently present in the top 10 genera or OTUs across studies [Figure 6]. Changes in *Bacteroides*, *Ruminococcaceae*, *Ruminococcus*, and *Roseburia* were consistently found to be in the top 10 most important variables across the different studies for both individuals with adenoma and carcinoma [Figure 6]. These data suggest that whether the non-resident bacterium is *Fusobacteria* or *Peptostreptococcus* may not be as important as how these bacteria interact with the changing resident community. Based on these observations, it is possible to hypothesize that small changes in community structure lead to new niches in which any one of the mouth-associated genera can gain a foothold, exacerbating the initial changes in community and facilitating the transition from adenoma to carcinoma stage of disease.

@@

One important caveat to this study is that even though genera associated with certain species such as *Bacteroides fragilis* and *Streptococcus gallolyticus* subsp. *gallolyticus* were not identified, it does not necessarily mean that these specific species are not important in human CRC [@bfrag_Boleij_2014; @strep_Aymeric_2017]. Since we are limited in our aggregation of the data to the genus level, it is not possible to clearly delineate which species are contributing to overall disease progression. Our observations are not inconsistent with the previous literature on either *Bacteroides fragilis* or *Streptococcus gallolyticus* subsp. *gallolyticus*. As an example, the stool-based full community models consistently identified the genus *Bacteroides*, as well as OTUs that classified as *Bacteroides*, to be important model components across studies. This suggests that even though *Bacteroides* may not increase the OR of individuals having an adenoma or carcinoma and may not vary in relative abundance, like *Fusobacterium*, it is still important in CRC. Additionally, *Streptococcus gallolyticus* subsp. *gallolyticus* is a mouth-associated microbe, and the results from this study suggest that regardless of sample type, mouth-associated genera are commonly associated with an increased OR for individuals to have either an adenoma or carcinoma.

The associations between the microbiota and indivdials with adenoma tumors are inconclusive, in part, because many studies may not be powered effectively to observe small effect sizes. None of the studies analyzed were properly powered to detect a 10% or lower change between cases and controls. The results within our meta-analysis suggest that a small effect size may well be the scope in which differences consistently occur between controls and those with adenomas. Future studies investigating adenoma tumors and the microbiota need to take power into consideration to reproducibly study whether the microbiota contributes to polyp formation. In contrast to adenoma stage of disease, our observations suggest that most studies analyzed have sufficient power to detect many changes in the carcinoma-associated microbiota because of large effect size differences between cases and controls [Figure 7].

### 2018-02-10 submission/manuscript.tex
update and improve discussion

Our study identifies clear differences in diversity, both at the
community level and for individual genera, present in patients with and
without CRC {[}Figure 1-3{]}. Although there was a step-wise decrease in
diversity as disease progressed from control to adenoma to carcinoma,
this did not translate into large effect sizes for the OR of tumors.
Even though mouth-associated genera increased the OR of having a
carcinoma, they did not consistently increase the RR of having an
adenoma. Additionally, our observations suggest that by combining
mouth-associated and CRC protective microbes we can classify either
adenoma or carcinoma stage of disease as well as models that use the
full community.

The data presented herein support the importance of select genera for
carcinoma, but not necessarily adenoma, formation. The results that we
have presented show that both the genera and OTU select and full models,
for the carcinoma stage of CRC, had similar AUCs {[}Figure 4 \& 5{]}.
This suggests that an interplay between a select number of potentially
protective and exacerbating microbes within the GI community is crucial
for carcinoma formation. Importantly, it suggests that there may be key
members of the GI community that might be studied further to potentially
reduce the risk of carcinoma. Conversely, using the present data, it is
clear that new approaches may be needed to identify members of the
community associated with adenoma stage of disease. Regardless of sample
type and whether a full or select model was used, our Random Forest
models consistently performed poorly. Yet, the step-wise decrease in
diversity suggests that the adenoma-associated community is not normal
but has changed subtly {[}Figure 1{]}. This change in diversity, at this
early stage of disease, could be focal to the adenoma itself. One
possible explanation is that how the host interacts with these subtle
changes at early stages of the disease is what leads to a thoroughly
dysfunctional community that is supportive of CRC genesis.

Within stool, common GI microbes were most consistently present in the
top 10 genera or OTUs across studies {[}Figure 6{]}. Changes in
\emph{Bacteroides}, \emph{Ruminococcaceae}, \emph{Ruminococcus}, and
\emph{Roseburia} were consistently found to be in the top 10 most
important variables across the different studies for both adenoma and
carcinoma {[}Figure 6{]}. These data suggest that whether the
non-resident bacterium is \emph{Fusobacteria} or
\emph{Peptostreptococcus} may not be as important as how these bacteria
interact with the changing resident community. Based on these
observations, it is possible to hypothesize that small changes in
community structure lead to new niches in which any one of the
mouth-associated genera can gain a foothold, exacerbating the initial
changes in community and facilitating the transition from adenoma to
carcinoma stage of disease.

@@

One important caveat to this study is that even though genera associated
with certain species such as \emph{Bacteroides fragilis} and
\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} were not
identified, it does not necessarily mean that these specific species are
not important in human CRC {[}20,22{]}. Since we are limited in our
aggregation of the data to the genus level, it is not possible to
clearly delineate which species are contributing to overall disease
progression. Our observations are not inconsistent with the previous
literature on either \emph{Bacteroides fragilis} or \emph{Streptococcus
gallolyticus} subsp. \emph{gallolyticus}. As an example, the stool-based
full community models consistently identified the genus
\emph{Bacteroides}, as well as OTUs that classified as
\emph{Bacteroides}, to be important model components across studies.
This suggests that even though \emph{Bacteroides} may not increase the
RR of CRC and may not vary in relative abundance, like
\emph{Fusobacterium}, it is still important in CRC. Additionally,
\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} is a
mouth-associated microbe, and the results from this study suggest that
regardless of sample type, mouth-associated genera are commonly
associated with an increased RR for both adenoma and carcinoma stage of
disease.

The associations between the microbiota and adenoma stage of disease are
inconclusive, in part, because many studies may not be powered
effectively to observe small effect sizes. None of the studies analyzed
were properly powered to detect a 10\% or lower change between cases and
controls. The results within our meta-analysis suggest that a small
effect size may well be the scope in which differences consistently
occur between controls and adenoma stage of disease. Future studies
investigating adenoma stage and the microbiota need to take power into
consideration to reproducibly study whether the microbiota contributes
to polyp formation. In contrast to adenoma stage of disease, our
observations suggest that most studies analyzed have sufficient power to
detect many changes in the carcinoma-associated microbiota because of
large effect size differences between cases and controls {[}Figure 7{]}.

>>>>

Our study identifies clear differences in diversity, both at the
community level and for individual genera, present in patients with
tumors versus controls {[}Figure 1-3{]}. Although there was a step-wise
decrease in diversity as disease progressed from control to adenoma to
carcinoma, this did not translate into large effect size increases in OR
for either adenoma or carcinoma tumors. Even though mouth-associated
genera increased individaul's OR of having a carcinoma, they did not
consistently increase the OR of having an adenoma. Additionally, our
observations suggest that by combining the top 5 most protective and
detrimental microbes we can classify indviduals with either adenoma or
carcinoma as well as models that use the full community.

The data presented herein support the importance of select genera for
carcinoma, but not necessarily adenoma, tumor formation. The results
that we have presented show that both the genera- and OTU-based select
and full models, for indviduals with carcinoma, had similar AUCs
{[}Figure 4 \& 5{]}. This suggests that an interplay between a select
number of potentially protective and exacerbating microbes within the GI
community could be crucial for carcinoma formation. Importantly, it
suggests that there may be key members of the GI community that should
be studied further to potentially help reduce the risk of carcinoma
tumor formation. Conversely, using the present data, it is clear that
new approaches may be needed to identify members of the community
associated with adenoma turmors. Regardless of sample type and whether a
full or select model was used, our Random Forest models consistently
performed poorly. Yet, the step-wise decrease in diversity suggests that
the adenoma-associated community is not normal but has changed subtly
{[}Figure 1{]}. This change in diversity, at this early stage of
disease, could be focal to the adenoma itself. Additionally, how the
host interacts with these subtle changes at early stages of the disease
could be what leads to a thoroughly dysfunctional community that is
supportive of tumorgenesis.

Within stool, common GI microbes were most consistently present in the
top 10 genera or OTUs across studies {[}Figure 6{]}. Changes in
\emph{Bacteroides}, \emph{Ruminococcaceae}, \emph{Ruminococcus}, and
\emph{Roseburia} were consistently found to be in the top 10 most
important variables across the different studies for both individuals
with adenoma and carcinoma {[}Figure 6{]}. These data suggest that
whether the non-resident bacterium is \emph{Fusobacteria} or
\emph{Peptostreptococcus} may not be as important as how these bacteria
interact with the changing resident community. Based on these
observations, it is possible to hypothesize that small changes in
community structure lead to new niches in which any one of the
mouth-associated genera can gain a foothold, exacerbating the initial
changes in community and facilitating the transition from adenoma to
carcinoma stage of disease.

@@

One important caveat to this study is that even though genera associated
with certain species such as \emph{Bacteroides fragilis} and
\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} were not
identified, it does not necessarily mean that these specific species are
not important in human CRC {[}20,22{]}. Since we are limited in our
aggregation of the data to the genus level, it is not possible to
clearly delineate which species are contributing to overall disease
progression. Our observations are not inconsistent with the previous
literature on either \emph{Bacteroides fragilis} or \emph{Streptococcus
gallolyticus} subsp. \emph{gallolyticus}. As an example, the stool-based
full community models consistently identified the genus
\emph{Bacteroides}, as well as OTUs that classified as
\emph{Bacteroides}, to be important model components across studies.
This suggests that even though \emph{Bacteroides} may not increase the
OR of individuals having an adenoma or carcinoma and may not vary in
relative abundance, like \emph{Fusobacterium}, it is still important in
CRC. Additionally, \emph{Streptococcus gallolyticus} subsp.
\emph{gallolyticus} is a mouth-associated microbe, and the results from
this study suggest that regardless of sample type, mouth-associated
genera are commonly associated with an increased OR for individuals to
have either an adenoma or carcinoma.

The associations between the microbiota and indivdials with adenoma
tumors are inconclusive, in part, because many studies may not be
powered effectively to observe small effect sizes. None of the studies
analyzed were properly powered to detect a 10\% or lower change between
cases and controls. The results within our meta-analysis suggest that a
small effect size may well be the scope in which differences
consistently occur between controls and those with adenomas. Future
studies investigating adenoma tumors and the microbiota need to take
power into consideration to reproducibly study whether the microbiota
contributes to polyp formation. In contrast to adenoma stage of disease,
our observations suggest that most studies analyzed have sufficient
power to detect many changes in the carcinoma-associated microbiota
because of large effect size differences between cases and controls
{[}Figure 7{]}.
